The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
Revisao da Secao 2.1: figura 4 separada em duas

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} to an internal event is only 
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements})~\cite{esterel.schizo}.
%
The example in the left of Figure~\ref{lst.delay} (\code{CODE-1}) relies on 
this restriction to avoid infinite execution:
Although both sides of the \code{par/or} have an \code{await} statement, the 
\code{loop} would restart instantaneously if the \code{await} could awake in 
the same reaction that reaches it.
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or making some transformation in the code to 
execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically, \emph{including the current reaction}, starting from requests of 
an internal event, as illustrated in the middle of Figure~\ref{lst.delay} 
(\code{CODE-2}).
In this case, moving the \code{await} to the end of the loop suffices to 
execute the periodic code immediately (and then in reaction to each \code{emit} 
request).
If the periodic \code{emit} depends on a condition, the code becomes more 
intricate, requiring to copy the condition test to the periodic code, as 
illustrated in the right of Figure~\ref{lst.delay} (\code{CODE-3}).

\begin{figure}
\begin{minipage}[t]{0.24\linewidth}
\begin{lstlisting}
event void e,f;
loop do
  par/or do
    await e;
  with
    emit e;
    await f;
  end
end

/* CODE-1 */
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.36\linewidth}
\begin{lstlisting}
event void e;
par do
  loop do
    <...>
    emit e;  // periodic request
  end
with
  loop do
    <...>    // code to execute
    await e; // await after
  end
end

/* CODE-2 */
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.36\linewidth}
\begin{lstlisting}
event void e;
var bool execute_now = false;
par do
  loop do
    <...>
    if <...> then
        execute_now = true;
        emit e;
    end
  end
with
  loop do
    if execute_now then
      <...> // code to execute
    end
    await e;
  end
end

/* CODE-3 */
\end{lstlisting}
\end{minipage}
%
\caption{ In \code{CODE-1}, the \code{await} occurs simultaneously with the 
\code{emit} and does not awake, preventing the \code{par/or} to terminate and 
execute the \code{loop} indefinitely.
\code{CODE-2} and \code{CODE-3} illustrate how to circumvent the \emph{delayed 
await} restriction.
\label{lst.delay}
}
\end{figure}

>>>>

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} to an internal event is only 
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements})~\cite{esterel.schizo}.
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // awakes 1s trail and restarts the loop
        await f;
    end
end
\end{lstlisting}

Even though both sides of the \code{par/or} have an \code{await} statement to 
avoid instantaneous termination, the \code{loop} would restart instantaneously 
if the \code{await} could awake in the same reaction that reaches it.

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or making transformations in the code to 
execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop makes the periodic 
code to also execute immediately (and then in reactions to each \code{emit} 
request).
If the periodic \code{emit} depends on a condition, the code becomes more 
intricate because we need to copy the condition test to the periodic code, as 
illustrated in the right of Figure~\ref{lst.delay}.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
event void e;
par do
    loop do
        <...>
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute
        await e;  // await after
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
event void e;
var bool should_execute = false;
par do
    loop do
        <...>
        if <...> then
            should_execute = true;
            emit e;
        end
    end
with
    loop do
        if should_execute then
            <...> // code to execute
        end
        await e;
    end
end
\end{lstlisting}
\end{minipage}
%
\caption{ Examples that circumvent the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop} (in the left), and by 
copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
